---
title: "viz_and_eda"
author: emanuel clemente
output: github_document
date: "2026-10-08"
---

# Global Settings for Every Chunk in the File

```{r setup, include=FALSE}

# include=FALSE (in the chunk header) runs this chunk but hides it from the knitted output.
# `echo = TRUE` shows each chunk's code above its output.

knitr::opts_chunk$set(echo = TRUE)
```

```{r}

# The function `library()` loads a package so its functions are available for the rest of the file.

library(p8105.datasets) # course datasets, including `weather_df`
library(tidyverse) # this package includes the functions `dplyr`, `ggplot2`, `tidyr`, `readr`, and more
library(haven) # this package reads SAS, SPSS, and Stata files
```

# Loading the Dataset

```{r}
data("weather_df")

weather_df =    # = at the start saves the result under this name.
  weather_df |> # the pipe `|>` means "and then": it passes the result on the left into the function on the right.
                # every line in a pipeline except the last ends with `|>`.
  mutate(       # the function `mutate()` adds or changes columns
    month = lubridate::floor_date(date, unit = "month") # `floor_date()` rounds each date down to the 1st of its month (example: 2021-03-17 -> 2021-03-01).
                                                        # `::` uses a function from a package without loading the whole package.
  )
```

```{r}
# tmax/tmin are in degrees C. prcp is in tenths of a mm (100 = 10 mm).

weather_df |> 
  ggplot(aes(x = prcp)) + # the function `aes()` maps columns to parts of the plot.
  geom_histogram()        # `ggplot` layers are joined with `+`, not `|>`.

# NOTE: messages about "bins = 30" or "removed rows" aren't errors; the removed rows are days with missing prcp.
```

```{r}

# The function `filters()` keeps only rows where the condition is TRUE.

weather_df |> 
  filter(prcp >= 1000) # 1000 tenths of a mm = 100 mm (about 4 inches) of rain in one day.
```

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |>                             # a comma between conditions means "and".
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) + # color and shape = name give each station its own color and point shape.
  geom_point(alpha = .75)                                       # alpha = transparency (0 = invisible, 1 = solid); helps when points overlap.
```

## `group_by()`

```{r}

# add some groups! The function `group_by()` tells R to treat each group separately in the steps that follow. By itself, it doesn't change the data; the header just shows "Groups: name, month, [72]"
# (3 stations x 24 months).

weather_df |>
  group_by(name, month)
```

## `summarize()`

```{r}

# the function `summarize()` collapses each group into ONE row of summary numbers.

weather_df |>
  group_by(name, month) |>
  summarize(
    count = n(),                # `n()` counts rows in each group
    n_days = n_distinct(date))  # `n_distinct()` countes unique values

# if `count` and `n_days` didn't match, there'd be duplicate rows.
# NOTE: "`summarize()` has grouped output by 'name'" is a message, not an error.
# R drops the last grouping variable (month) after summarizing.

```

DON'T DO THIS

```{r}
weather_df |>
  pull(tmax) |> # the function `pull()` pulls one column out as a plain vector.
  summary()     # the function `summary()` describes that column.

# avoid doing this format as it lumps all three stations together, and the output isn't a data frame, so you can't keep piping it.
```

Do this instead!

```{r}
weather_df |>
  group_by(name) |>                                 # one summary row per station
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),           # `na.rm = TRUE` skips missing values ('NA') 
    median_tmax = median(tmax, na.rm = TRUE),
    q95_prcp = quantile(prcp, 0.95, na.rm = TRUE)   # 95th percentile of daily rain
  ) |>
  knitr::kable(digits = 2)                          # the function `kable()` makes a formatted table in the knitted file. `digits = 2` rounds to 2 decimals.
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |>
  # data is "long" here (one row per station-month).
  pivot_wider(
    names_from = name,        # each station becomes its own column
    values_from = mean_tmax   # wide is easier to read, but keep data long for analysis and plotting.
  ) |>
  knitr::kable(digits = 2)
```

```{r}
weather_df |>
  group_by(name, month) |>
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |>
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +         # a dot for each monthly average
  geom_line()            # connects the dots, with a separate line for each station
```

```{r}
weather_df |>
  group_by(name) |>
  mutate(center_tmax = tmax - mean(tmax, na.rm = TRUE)) |>   # the function `mutate()` keeps every row; `summarize()` collapses them.
  ggplot(aes(x = date, y = center_tmax, color = name)) +     # when data is grouped, `mean()` inside `mutate()` is calculated within each group.
                                                             # how much hotter or colder each day was than that STATION's own average.
  geom_point()
```

## What About "Window" Functions?

```{r}

# Window functions return one value per row, so they go in `mutate()`, not `summarize()`.

weather_df |>
  group_by(name, month) |>
  mutate(temp_rank = min_rank(desc(tmax))) |> # the function `min_rank()` ranks smallest = 1. `desc()` flips it so the HOTTEST day = 1.
  filter(temp_rank < 2)                       # keeps the hottest day per station per month; ties are all kept.
```

lead and lag

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    lagged_tmax = lag(tmax),   # value from the previous row (yesterday)
    lead_tmax = lead(tmax)     # value from the next row (tomorrow)
  )

# the function `group_by(name) keeps each station separate, so the first day of one station.
# doesn't borrow a value from the last day of another; it gets 'NA' instead.
# lag/lead follow the current row order. With unsorted data, add `arrange(date)` after `group_by()`.
```

```{r}
weather_df |>
  group_by(name) |>
  mutate(
    lagged_tmax = lag(tmax),
    temp_change = tmax - lagged_tmax
  ) |>
  summarize(
    mean_temp_change = mean(temp_change, na.rm = TRUE),  # average daily change (close to 0)
    sd_temp_change = sd(temp_change, na.rm = TRUE)       # bigger SD = bigger day-to-day swings
  )

# common pattern: build the variable you need with `mutate()`, then `summarize()` it.
```

## Revisiting Some Examples

import, clean, tidy, etc. the pulse data, and compute mean and median BDI score at each visit.

```{r}
pulse_df =
  haven::read_sas("data/public_pulse_data.sas7bdat") |> # reads the SAS file
  janitor::clean_names() |> # the function `clean_names()` makes names lowercase with underscores (example: BDI_score_BL -> bdi_score_bl)
  pivot_longer(
    bdi_score_bl:bdi_score_12m,        # columns to stack; `:` means "from this value through that value".
    names_to = "visit",                # old column names go into a new "visit" column.
    names_prefix = "bdi_score_",       # strips this text, so "bdi_score_bl" -> "bl".
    values_to = "bdi"                  # the scores go into a new "bdi" column.
  ) |>  # wide (one row per person) -> long (one row per person per visit). Long is the tidy format.
  select(id, visit, everything()) |>   # the function `select()` picks and orders columns; `everything()` = all remaining columns.
  mutate(
    visit = replace(visit, visit == "bl", "00m") # replace(column, condition, new value). == tests equality.
  )                                              # baseline becomes "00m" so visits sort in time order: 00m, 01m, 06m, 12m.

pulse_df |>
  group_by(visit) |>
  summarize(
    mean_bdi = mean(bdi, na.rm = TRUE),
    median_bdi = median(bdi, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 2)
```

in the FAS data, compute mean outcome (ears only) across dose and day of treatment; show in a reader-friendly table.

```{r}
# pups file: one row per pup; litter fileL one row per litter, with treatment group.

pups_df =
  read_csv("data/FAS_pups.csv", skip = 3, na = c("", ".", "NA")) |> # `skip = 3` ignores the first 3 lines of the file (for notes above the column names).
                                                                    # `na = c(...)` lists every way the file marks a missing value.
  janitor::clean_names()

litters_df =
  read_csv("data/FAS_litters.csv", na = c("", ".", "NA")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), 3)     # `c()` combines values into a vector.
                                                        # the 3 splits after the 3rd character: "Con7" -> dose "Con", day_of_tx "7".

fas_df =
  left_join(pups_df, litters_df, by = "litter_number") |> # keeps every pup row and adds matching litter info, matched on `litter_number`; pups with no matching litter get `NA`.
  select(litter_number, dose, day_of_tx, everything()) |>
  drop_na(dose, day_of_tx)    # removes rows with. `NA` in these columns (pups with no treatment group).

fas_df |>
  group_by(dose, day_of_tx) |>
  summarize(
    mean_ears = mean(pd_ears, na.rm = TRUE) # `pd_ears`= postnatal day the pup's ears unfolded
  ) |>
  pivot_wider(
    names_from = day_of_tx, # one column per treatment day
    values_from = mean_ears
  ) |>
  knitr::kable(digits = 2)  # result: one row per dose, one column per day
```